---
title: "Content 1-2: Answers to practice questions"
date: "Monday, September 25, 2023"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_depth: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains answers to the practice questions in Content 2-1. 

In order for the code in this document to work, we must still load the `tidyverse` family of packages and load our dataset.
```{r}
library(tidyverse)
load("nmes2018.rda")
```

# Filter rows with `filter()`

## Practice

1. Find all individuals who meet the following characteristics:

a. Had more than $1,000 but less than $10,000 in medical expenditures: *There are 1,131 of these individuals.*
```{r}
filter(nmes.data, totalexp > 1000, totalexp < 10000)
```

b. Always wear a seatbelt: *There are 2,283 of these individuals.*
```{r}
filter(nmes.data, beltuse == "Always")
```

c. Have a BMI of at least 30 and currently smoke: *There are 148 of these individuals.*
```{r}
filter(nmes.data, bmi >= 30, current == "Yes")
```

d. Are both poor and college grads: *There are 58 of these individuals.*
```{r}
filter(nmes.data, poor == "Poor", educate == "CollGrad")
```

e. Quit smoking more than 10 years ago: *There are 439 of these individuals.*
```{r}
filter(nmes.data, yearsince > 10)  
```

f. Quit smoking more than 10 years ago and are younger than 25 years old: *There are 0 of these individuals.*
```{r}
filter(nmes.data, yearsince > 10, age < 25)     
```

g. Quit smoking more than 10 years ago or are younger than 25 years old: *There are 852 of these individuals.*
```{r}
filter(nmes.data, yearsince > 10 | age < 25)       
```

2. Another useful filtering helper is `between()`.  Look at the help file for this function using `?between`.  Can you use it to simplify the code below?
```{r}
filter(nmes.data, totalexp >= 1000, totalexp <= 10000)
filter(nmes.data, between(totalexp, 1000, 10000))
```

3. How many individuals are missing `bmi` values?

There are 124 individuals missing `bmi` values:
```{r}
no_bmi <- filter(nmes.data, is.na(bmi))
nrow(no_bmi)
```

4. How many individuals are missing `current` values?  Filter to these individuals.  Do you think this is really *missing* data?

There are 2,084 individuals missing current values. These individuals all have `no` values for `eversmk`, so it looks like these individuals have never been smokers.  So it's not really missing data, we know these are not current smokers!
```{r}
no_current <- filter(nmes.data, is.na(current)) 
nrow(no_current)

no_current
```


# Arrange rows with `arrange()`

## Practice

5. Sort individuals in this dataset to find the youngest people. Remember you can force `R` to show all the columns with the `print()` function and `width = Inf`.

To see the age variable, you might need to increase your window size, since it is one of the later variables.  Alternatively, you could use the `print()` function with `width = Inf` to force all the columns to be shown. We can see the youngest individuals are 19 years old.
```{r}
arrange(nmes.data, age)

print(arrange(nmes.data, age), width = Inf)
```

6. Sort individuals in this dataset to find the oldest never smokers.

The first category for `eversmk` is `no`, so the never smokers will be first if we sort by `eversmk`.  Then we want to sort by `desc(age)` to get the oldest ages first within the smoking categories.  We see the oldest never smokers are 94 years old.
```{r}
levels(nmes.data$eversmk)

arrange(nmes.data, eversmk, desc(age))

print(arrange(nmes.data, eversmk, desc(age)), width = Inf)
```

7. Sort individuals in this dataset to find the oldest ever smokers

Now we want to sort first by `desc(eversmk)` to put the ever smokers first.  Then sort by `desc(age)` to get the oldest individuals within the smoking groups first.  The oldest ever smokers are also 94 years old.
```{r}
arrange(nmes.data, desc(eversmk), desc(age))

print(arrange(nmes.data, desc(eversmk), desc(age)), width = Inf)
```

8. Sort individuals in this dataset to find the never smokers with the highest medical expenditures.

Sort first by `eversmk` to get the never smokers first; then by `desc(totalexp)` to get the highest expenditures first.  The highest expenditure among the nevers smokers is $128,157.
```{r}
arrange(nmes.data, eversmk, desc(totalexp))
```


# Select columns with `select()`

## Practice

9. What happens if you include the name of a variable multiple times in a `select()` call?

You still just get each variable once, not multiple copies of the same variable.

```{r}
select(nmes.data, age, ageCat, bmi, bmi)
```

10. What does the `any_of()` function do? Run the code below to figure it out:

It allows you to select any variables whose name is included in the list of variables you specified and stored as `myVars`.  Selecting variables this way can be helpful when you want to be able to change the list of variable names you want to select in an organized way.
```{r}
myVars <- c("totalexp", "mscd", "age", "eversmk", "bmi")
select(nmes.data, any_of(myVars))
```


# Add new variables with `mutate()`

## Practice

11. Create the following new variable and add to our NMES dataset in a single mutate command:

* Variable that tells whether an individual always uses their seat belt
* Variable that tells whether an individual's medical expenditures are higher than the mean expenditure value in the dataset
* Variable that tells whether an individual quit smoking more than 5 years ago

```{r}
nmes.data.2 <- mutate(nmes.data, 
                      always_belt = (beltuse == "Always"),
                      exp_above_mean = (totalexp > mean(totalexp)),
                      former_5 = yearsince > 5)

print(nmes.data.2, width = Inf)
```