---
title: "Content 3-1: Answers to practice questions"
date: "Monday, October 9, 2023"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_depth: 1
---
```{css, echo=FALSE}
.scroll-200 {
  max-height: 200px;
  overflow-y: auto;
  background-color: inherit;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This document contains answers to the practice questions in Content 3-1. 

In order for the code in this document to work, we must still load the `tidyverse` family of packages and load our dataset.
```{r}
library(tidyverse)
load("nmes2018.rda")
```

# Working with data tibbles and data frames

## Subsetting data sets to individual variables

## Practice

1. What's the difference between the results of these two sets of code?

```{r eval=FALSE}
nmes.data %>%
  .$age

nmes.data %>%
  select(age)
```

**The first one returns a vector with all the age values in it.  The second one returns a data frame that has one variable in it -- age.**

2. If you have the name of a variable stored in an object like `var <- "age"`, how can you extract that variable from a tibble?  Which one of these works?  Why?

```{r eval=FALSE}
var <- "age"

nmes.data %>%
  .$var

nmes.data %>%
  .[[var]]
```

**Only the second code options works in this case.  That's because only the second option, the square brackets, allows us to refer to a variable name in quotes.**

# Importing data using functions from `readr`

## Using `read_csv()` to import data

## Practice

3. What to the functions `read_tsv()` and `read_delim()` do? 
```{r}
?read_tsv
?read_delim
```

**`read_tsv()` reads in data where the values are separated by tabs instead of commas.  `read_delim()` reads in data were the values are separated by a specified delimiter.  You can specify the particular delimiter you are using with the `delim = ` option.**

## Parsing a file

## Writing to a file

