---
title: "Content 1-2: Answers to practice questions"
date: "Monday, September 18, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains answers to the practice questions in Content 1-2. 

In order for the code in this document to work, we must still load the `tidyverse` family of packages and load our dataset.
```{r}
library(tidyverse)
load("nmesSub.rda")
```

### Bar charts

#### Practice

1. Make a bar chart that shows the proportion of individuals in our data set with and without a major smoking-caused disease (`mscd`).
```{r}
ggplot(data = nmes.data, mapping = aes(group = 1)) + 
  geom_bar(mapping = aes(x = mscd, y = stat(prop)))
```

### Position adjustments

#### Practice

2. Make a bar chart that allows you to compare the proportion of individuals with a major smoking-caused disease (`mscd`) between those who have and haven't ever smoked (`eversmk`).

Since we want to know the proportion of major smoking-caused disease between smoker and non-smokers, we need to know the proportion within the `eversmk` groups.  The easiest way to do this is with the `"fill"` position.  
```{r}
ggplot(data = nmes.data) + 
  geom_bar(mapping = aes(x = eversmk, y = stat(count), fill = mscd), position = "fill")
```

If you switch the `x` and `fill` aesthetics, the plot isn't quite right -- it lets us compare the proportion of smokers between the two disease status groups instead:
```{r}
ggplot(data = nmes.data) + 
  geom_bar(mapping = aes(x = mscd, y = stat(count), fill = eversmk), position = "fill")
```

### Coordinate systems

#### Practice

3. Flip the coordinates on your bar chart that compares the proportion of individuals with a major smoking-caused disease (`mscd`) between those who have and haven't ever smoked (`eversmk`).
```{r}
ggplot(data = nmes.data) + 
  geom_bar(mapping = aes(x = eversmk, y = stat(count), fill = mscd), position = "fill") + 
  coord_flip()
```

4. Create a zoomed-in scatterplot of `bmi` vs. `age` with only bmi values between 20-25 and only ages between 40-60.
```{r}
ggplot(data = nmes.data) +
  geom_point(mapping = aes(x = age, y = bmi)) +
  coord_cartesian(xlim = c(40, 60), ylim = c(20, 25))
```

### Pulling it all together -- the layered language of graphics

### Boxplots

#### Practice

5. Make a graphic that shows the relationship between total medical expenditures (`totalexp`) and smoking (`eversmk`) separately for the three seat-belt groups.
```{r}
ggplot(data = nmes.data) +
  geom_boxplot(mapping = aes(x = eversmk, y = totalexp)) + 
  facet_wrap(~ beltuse)
```

6. Make a graphic that shows the relationship between total medical expenditures (`totalexp`) and smoking (`eversmk`) separately for the three seat-belt groups, where the boxplots are colored differently for the different smoking (`eversmk`) groups.
```{r}
ggplot(data = nmes.data) +
  geom_boxplot(mapping = aes(x = eversmk, y = totalexp, color = eversmk)) + 
  facet_wrap(~ beltuse)
```

```{r}
ggplot(data = nmes.data) +
  geom_boxplot(mapping = aes(x = eversmk, y = totalexp, fill = eversmk)) + 
  facet_wrap(~ beltuse)
```

7. Flip the coordinates of the graph you made in 6.
```{r}
ggplot(data = nmes.data) +
  geom_boxplot(mapping = aes(x = eversmk, y = totalexp, fill = eversmk)) + 
  facet_wrap(~ beltuse) + 
  coord_flip()
```

8. If you type `?geom_boxplot` in the console window to open the help file for this geom, you'll see there is an option called `varwidth` with a default value of `FALSE`.  What happens if you set this option to `TRUE`?  Create the two plots with below, and also scroll down through the help file to see what it says about `varwidth`.
```{r}
ggplot(data = nmes.data) +
  geom_boxplot(mapping = aes(x = mscd, y = totalexp), varwidth = FALSE)

ggplot(data = nmes.data) +
  geom_boxplot(mapping = aes(x = mscd, y = totalexp), varwidth = TRUE)
```

The first plot, with `varwidth = TRUE` has boxplots of the same width but the second plot, with `varwidth = FALSE` has a narrower boxplot for the "Yes" MSCD group compared to the "No" MSCD group.  From the help file: "`varwidth`: If `FALSE` (default) make a standard box plot. If `TRUE`, boxes are drawn with widths proportional to the square-roots of the number of observations in the groups (possibly weighted, using the weight aesthetic)."  So when this is set to `TRUE`, the width of the box reflects the number of observations included in the box.  This is helpful for comparing sample sizes across groups.