---
title: "Content 2-2: Answers to practice questions"
date: "Monday, October 2, 2023"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_depth: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains answers to the practice questions in Content 2-2. 

In order for the code in this document to work, we must still load the `tidyverse` family of packages and load our dataset.
```{r}
library(tidyverse)
load("nmes2018.rda")
```

# Grouped summaries with `summarize()` and `group_by()`

## Practice:

1. Calculate the mean medical expenditures for each of the age groups defined by `ageCat`.  What happens to medical expenditures with increasing age?
```{r}
by_age <- group_by(nmes.data, ageCat)

summarize(by_age,  mean_exp = mean(totalexp))
```

**With increasing age across these age groups, the mean medical expenditure also increases.**

2. Calculate the minimum and maximum ages for each of the education groups defined by `educate`.
```{r}
by_edu <- group_by(nmes.data, educate)

summarize(by_edu,  
          min_age = min(age),
          max_age = max(age))
```

**All of these education groups have similiar age ranges, except there is a higher minimum age among college graduates.  This makes sense, since college graduates are generally at least 21 years old.**

## Summarizing counts

## The `count()` function

## Grouping by multiple variables

## Practice

3. Find the number of people, mean, and median medical expenditures for those who have and don't have a major smoking-caused disease (`mscd`).
```{r}
by_mscd <- group_by(nmes.data, mscd)

summarize(by_mscd, 
          n = n(),
          mean_exp = mean(totalexp),
          median_exp = median(totalexp))
```

4. Find the number of people, mean, and median BMIs for those who have and haven't ever smoked (`eversmk`).

Here we need to use `sum(!is.na())` to count the number of individuals, since there is missing BMI data.  We also need to add the `na.rm = TRUE` option for our summary statistics.
```{r}
by_smk <- group_by(nmes.data, eversmk)

summarize(by_smk, 
          n = sum(!is.na(bmi)),
          mean_exp = mean(bmi, na.rm = TRUE),
          median_exp = median(bmi, na.rm = TRUE))
```

5. Find the number of people and the minimum and maximum medical expenditures for each combination of poverty status (`poor`) and education level (`educate`).
```{r}
by_poor_educate <- group_by(nmes.data, poor, educate)

summarize(by_poor_educate, 
          n = n(),
          min_exp = min(totalexp),
          max_exp = max(totalexp))
```

# Combining multiple operations with the pipe, `%>%`

## Practice

6. Use the pipe to perform the following sequence of operations on the NMES data: mutate to create a log10 medical expenditures variable, filter to only people older than 65, group by smoking status, summarize to get median log10 expenditures and the number of people in each smoking group
```{r}
nmes.data %>%
  mutate(log_exp = log10(totalexp)) %>%
  filter(age > 65) %>%
  group_by(eversmk) %>%
  summarize(n = n(), med_log_exp = median(log_exp))
```

## Back to grouping by multiple variables

## Practice

7. Use the pipe to calculate the proportions of people with and without MSCD within each smoking group.
```{r}
nmes.data %>%
  group_by(eversmk, mscd) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n))
```

**The proportion with MSCD is higher among smokers compared to non-smokers.**

8. Modify your code from 7 to only give the proportions *with* MSCD for each smoking group.

**Here we need to filter our results to only choose the rows with `mscd == "Yes"`:**
```{r}
nmes.data %>%
  group_by(eversmk, mscd) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n)) %>%
  filter(mscd == "Yes")
```

**Note we could clean up this table of proportions even more by removing the `mscd` column by adding a `select()` function at the end.**  I also added a more informative name to the proportion column to get a nice table showing the risk of MSCD within each smoking group.
```{r}
nmes.data %>%
  group_by(eversmk, mscd) %>%
  summarize(n = n()) %>%
  mutate(risk_of_mscd = n/sum(n)) %>%
  filter(mscd == "Yes") %>%
  select(-mscd)
```

## Ungrouping

# Combining `dplyr` with `ggplot2`

## Practice

9. Use the pipe to create a scatterplot of log10 medical expenditures vs. age, with points colored differently for smokers and non-smokers.  Add smoothed lines for each smoking group as well.
```{r}
nmes.data %>%
  mutate(log_exp = log10(totalexp)) %>%
  ggplot(mapping = aes(x = age, y = log_exp, color = eversmk)) +
  geom_point() + 
  geom_smooth(se = FALSE)
```